Precise, Second-Order Verification

### 2014-11-13 slides/nus-talk.tex
NUS slides

@@

\begin{frame}[fragile]{Motivation}
Does this program terminate?
\begin{center}
\begin{minipage}{0.4\linewidth}
 \begin{lstlisting}[language=C,basicstyle=\normalsize]
while (x > 0) {
  x++;
}
 \end{lstlisting}
\end{minipage}
\end{center}

That depends...

It \emph{does not} terminate if the program operates on integers...

But it \emph{does} terminate if the program operates on fixed-width words
(-x decreases monotonically and is bounded by -MAXINT).

\begin{frame}[fragile]{Motivation}
Which of these guys terminates?
\begin{center}
\begin{minipage}{0.45\linewidth}
 \begin{lstlisting}[language=C,basicstyle=\normalsize]
while (x > 0) {
  x = (x - 1) & x;
}
\end{lstlisting}
\end{minipage}
\begin{minipage}{0.45\linewidth}
 \begin{lstlisting}[language=C,basicstyle=\normalsize]
while (x > 0) {
  x = (x + 1) & x;
}
\end{lstlisting}
\end{minipage}
\end{center}

The left one does (x decreases monotonically), but the right one doesn't ($2 \rightarrow 2 \rightarrow \dots$).
\end{frame}

\begin{frame}{Motivation}
Proving program termination has been a big deal for a while.

It'd be nice if we could automatically prove termination for programs that run on computers.
\end{frame}

\section{Proving Termination}
\begin{frame}{What Does a Termination Proof Look Like?}
We prove that a program terminates by finding a \emph{ranking function}.

@@

We are therefore in the business of finding ranking functions.
\end{frame}

\begin{frame}{Finding Ranking Functions}
Traditionally, people find ranking functions under some assumptions:

\begin{itemize}
 \item Programs are linear
 \item Ranking functions are linear
 \item Program variables are rationals
\end{itemize}

With these assumptions in hand, one can deploy a very powerful piece of linear algebra: \emph{Farkas's lemma}.

This lemma makes light work of a ton of termination problems.

But it's not for us.
\end{frame}

@@

\section{Program Synthesis}
\iffalse
\begin{frame}{Concrete Synthesis in C}
 We're using C as an implementation language.  To do that we need:

@@

\section{Advanced Topics}
\begin{frame}{Conditional Termination}

>>>>

\begin{frame}{Existential Second-Order Logic}
 Second-order logic allows quantification over functions as well as ground terms.

 Existential second-order logic is a restriction where we can only existentially quantify over functions.

\begin{frame}{Existential Second-Order Logic Example}
The following is an existential second-order formula:

 \[
  \exists F . \forall x . F(x) > x
 \]

 When using Peano arithmetic as a background theory, this formula is satisfiable.  One satisfying assignment would be $F(x) = x + 1$.
\end{frame}

\begin{frame}{Second-Order SAT}
 A further restriction is to say that all the first-order variables are boolean propositions.  This
 gives us the \emph{second-order SAT problem}.

 Second-order SAT is an extension of propositional SAT, where we are allowed to quantify
 over functions.

 Second-order SAT is NEXPTIME-complete.
\end{frame}

\begin{frame}{Using Second-Order SAT}
 Second-order SAT is a very general problem and can be used for lots of problems, including:

 \begin{itemize}
  \item Termination and non-termination
  \item Safety invariants
  \item Finding bugs without unwinding loops
  \item Code refactoring
  \item Code deobfuscation
  \item Program synthesis
  \item Concurrency
 \end{itemize}

 Also since everything boils down to SAT, we can freely combine with any theory that is decidable in NP.
\end{frame}

\begin{frame}{Solving Second-Order SAT}
 To solve second-order SAT instances, we use a program synthesiser.

 The synthesiser finds programs computing each of the quantified functions.
\end{frame}

\begin{frame}[fragile]{Example}
Consider the following second-order SAT problem:

 \[
  \exists F . \forall x . F(x) < x
 \]

This formula is satisfiable.  To show this, our program synthesiser might generate the following program:

\begin{center}
\begin{lstlisting}
F(unsigned int x) {
  unsigned int t1 = x / 2;
  return t1;
}
\end{lstlisting}
\end{center}
\end{frame}

@@

\begin{frame}{What Does a Termination Proof Look Like?}
We prove that a program terminates by finding a \emph{ranking function}.

@@

We are therefore in the business of finding ranking functions.
\end{frame}
% 
% \begin{frame}{Finding Ranking Functions}
% Traditionally, people find ranking functions under some assumptions:
% 
% \begin{itemize}
%  \item Programs are linear
%  \item Ranking functions are linear
%  \item Program variables are rationals
% \end{itemize}
% 
% \vspace{1em}
% 
% With these assumptions in hand, one can deploy a very powerful piece of linear algebra: \emph{Farkas's lemma}.
% 
% This lemma makes light work of a ton of termination problems.
% 
% \vspace{1em}
% 
% But it's not for us.
% \end{frame}

@@

% \section{Program Synthesis}
\iffalse
\begin{frame}{Concrete Synthesis in C}
 We're using C as an implementation language.  To do that we need:

@@

% \section{Advanced Topics}
\begin{frame}{Conditional Termination}
